---
title: "Territorial configuration of Colombian provinces"
subtitle: 'Measures of rurality and connectivity'
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      collapse=TRUE,
                      fig.align="center")
```

## 1. Background project

This project contains the methods, algorithms and visualizations employed by me and Dr. Juan Carlos Ramírez in the technical report *Territorial configuration of Colombian provinces* (original title in Spanish), published by ECLAC (Economic Commission for Latin America and the Caribbean).

A first version of such document was published in 2017, and can be at ECLAC's [digital repository](https://repositorio.cepal.org/handle/11362/40852). An actualization is to be published in 2022. 

## 2. Motivation

This study aims to characterize and quantify a rurality measure for Colombian provinces, and to use it to explore the rural-urban linkages and the centrality measures of these territorial entities. The Provincial Demographic Urbanization Index (IDUP, in Spanish) is constructed via Principal Component Analysis (PCA). It derives into three groups: urban, intermediate and rural. Ten refined categories of rurality are defined combining the IDUP and the travel times by car between the largest population seats.

For each province, its rurality category is an intrinsic function of its demographic characteristics and its relationship with the most nearby urban center. This single connection implies a reduced vision of rural-urban relationships, which in general can encompass more complex entanglements of territorial, economic and social relations. The analysis of the complete network topology is necessary to define and explore provinces' centralities, and to estimate the impact of the influence of other key territories. In total, 751 rural-urban linkages are detected amongst Colombian provinces. The strength of each link is a function of differences in the IDUP values and the spatial distance between connected provinces.

In the network of rural-urban linkages different node centralities can approximate measures of degrees of connection, intermediation, and influence. This allow us to construct the categories of centrality, and distinguish central provinces, isolated provinces, influencial provinces, and 'bridge' provinces.

## 3. Creating an index for rurality at the province level
```{r cars}
# Packages to employ
library("dplyr")
library("ggplot2")
library("scales")

###########################################################################################
# INPUT DATA
###########################################################################################

# Demographic data on municipalities, from Colombia's census (2018), projected to 2021 
demo_m = read.csv2('data/population_by_municipality.csv',
                  header = TRUE, stringsAsFactors = FALSE)
head(demo_m)

# Membership of every municipality to every province 
m_to_p = read.csv2('data/municipality_to_province.csv',
                  header = TRUE, stringsAsFactors = FALSE)  
head(m_to_p)  

###########################################################################################
# AGGREGATED DATA
###########################################################################################

# Aggregate data to province level
datap = demo_m %>% 
  left_join(m_to_p, c('codmun'='codmun')) %>% # Join by municipality code
  group_by(codpro,Provincia) %>% summarise(
    # departamentos = paste0(unique(Departamento),collapse=", ")
    n_municipios = n_distinct(codmun), # number of municipalities
    pob_total = sum(Total), # Total population size
    pob_cabecera = sum(Cabecera), # Population size living in the municipal seat
    pob_disperso = sum(Disperso), # Population size living outside the municipal seat
    area = sum(Area), # Total area
    densidad_pob = sum(Total)/sum(Area), # Average population density
    imp_cabeceras = 100*sum(Cabecera)/sum(Total), # Avg. % of population living in the municipal seat
    max_cabeceras = max(Cabecera) # Population size on largest municipal seat
  )
datap = datap %>% 
  left_join(demo_m %>% # Joint province-level data with largest municipal seat identification
              dplyr::select(codmun,Municipio,Cabecera),
            c('max_cabeceras'='Cabecera'))
head(datap)

# Data to apply Principal Component Analysis on: log(population density) + 
# Avg. % of population living in the municipal seat + log(size of largest municipal seat)
data_pc = datap %>% mutate(log_densidad_pob = log(densidad_pob),
                          log_max_cabeceras = log(max_cabeceras)) %>% 
  dplyr::select(log_densidad_pob,imp_cabeceras,log_max_cabeceras)
head(data_pc)

###########################################################################################
# CREATE AN INDEX WITH PRINCIPAL COMPONENTS ANALYSIS
###########################################################################################

# Applying Principal Component Analysis
pc = prcomp(data_pc[,-1], center = TRUE, scale. = TRUE)
summary(pc) # The first component explains 77.1% of total variance

# Standardized index:
# The Demographic Urbanization Index of Provinces
IDUP = 100*rescale(pc$x[,1])
datap$IDUP = IDUP

# This index is a linear combination of transform variables.
# Coefficients can be recovered with linear regression
coef(lm(IDUP~log(densidad_pob)+(imp_cabeceras)+log(max_cabeceras), datap))

#IDUP = -46.388 +4.128*log(densidad_pob) +0.356*imp_cabeceras +4.866*log(max_cabeceras) 

###########################################################################################
# CREATE THREE BROAD GROUPS OF RURALITY
###########################################################################################

# Using cutoff defined in previous methodology, 42--65, we divide provinces
# into three groups: urban, intermediate and rural
datap = datap %>% 
  mutate(grupo = case_when(IDUP > 65 ~ "Urbana",
                           IDUP < 42 ~ "Rural",
                           TRUE ~ "Intermedia"))

# Save the data frame
# write.csv2(datap %>% ungroup,'agg_provincias.csv')

# Histogram of The Demographic Urbanization Index of Provinces
datap %>% ggplot(aes(x=IDUP)) + 
  geom_histogram(fill="blue",bins = 40) +
  geom_vline(xintercept = 42, linetype="dashed", color = "red", size=1.5) +
  geom_vline(xintercept = 65, linetype="dashed", color = "red", size=1.5) + 
  annotate(geom="text", x=72, y=15, label="Urban ->", color="red") +
  annotate(geom="text", x=52, y=15, label="Intermediate ->", color="red") +
  annotate(geom="text", x=35, y=15, label="<- Rural", color="red") +
  theme_bw() + 
  labs(x='IDUP - Demographic Urbanization Index of Provinces',
       y='Count of provinces')
```

## 4. Extracting distance and time-to-travel between provinces using Google Maps

To be able to extract the distance and time-to-travel between two provinces in Colombia we employ the [Distance Matrix API service by Google Maps](https://developers.google.com/maps/documentation/javascript/distancematrix). This requires creating a Google Developers account, registering a valid credit card and requesting a private API key. *Do not worry!* We will not be charged if our data queries are below the limits:

- 2,500 free elements per day, calculated as the sum of client-side and server-side queries. 
- Maximum of 25 origins or 25 destinations per request. 
- 100 elements per request. 
- 100 elements per second, calculated as the sum of client-side and server-side queries.

*We will set the queries to respect these limits to avoid being charged* by constructing queries in a hierarchical and partitioned fashion. First, distances/times to the largest municipality of each urban province, which help us to create a further detailed set of rural/urban categories; then to the largest municipality of each intermediate province; and finally all the remaining pairwise connections.

```{r, eval = FALSE}
# Packages to employ
library('gmapsdistance')
library('stringi')

# Setting your private API key for Google Maps Distance Matrix API service
# set.api.key("XXX")

###########################################################################################
###########################################################################################
# DISTANCE AND TIME-TO-TRAVEL TO URBAN PROVINCES
###########################################################################################
###########################################################################################

###########################################################################################
# INPUT DATA
###########################################################################################

datag = datap[,c(1,12)] %>% # Provinces info (rurality group)
  left_join(datam[,c(2:4)],c('codmun'='codmun')) %>% # Municipalities info (name, geo: lat, long)
  mutate(origin = gsub(' ','+', stri_trans_general( # Clean accents on municipalities and department  names
    paste(Municipio,Departamento,'Colombia'),id = "Latin-ASCII") # Combined name: municipality, department and Colombia
  ))

urb = c('Bogota+Colombia',datag$origin[2:14]) # Names of largest municipality of each urban province

###########################################################################################
# MAKE OPTIMAL SIZED QUERIES (TIME TO TRAVEL TO URBAN PROVINCES)
###########################################################################################

# Since there are 14 urban provinces, we need 19 queries against 7 origins and a remaining query
# To keep query size below 100 elements
requ = list()
for(k in 1:19){
  requ[[k]] = gmapsdistance(datag$origin[(k*7+8):(k*7+14)], 
                            destination = urb, mode = "driving")
  print(k)
  Sys.sleep(60) # Make a query past one minute, to not overload the API
}

# Last query of remaining size
requ[[20]] = gmapsdistance(datag$origin[148:150], 
                           destination = urb, mode = "driving")

# Extract all the distances and time-to-travel from queries' replies
timeUrb = requ[[1]]$Time
distanceUrb = requ[[1]]$Distance
for(k in 2:20){
  timeUrb = rbind.data.frame(timeUrb, requ[[k]]$Time)
  distanceUrb = rbind.data.frame(distanceUrb, requ[[k]]$Distance)
}

# Save data
write.csv2(timeUrb,'time_urbanas.csv')
write.csv2(distanceUrb,'distance_urbanas.csv')

###########################################################################################
# FEED A MORE DETAILED SET OF URBAN/RURAL CATEGORIES
###########################################################################################

datag = datag %>% # Identify to which province each 'nearest' urban setting belongs
  left_join(timeUrb %>% melt() %>% group_by(or) %>% summarise(nearest = min(value,na.rm = T)) %>%
              left_join(timeUrb %>% melt(), c('or'='or','nearest'='value')), c('origin'='or'))

# Turn time-to-travel from being mesured in seconds to hours  
datap$nearest.time = datag$nearest/(60^2)  
# Extract the name of the 'capital' of nearest urban province
datap$nearest.urb = gsub('.{1}$', '',str_match(datag$variable, "Time.\\s*(.*?)\\s*+Colombia")[,2])

# More detailed categories of rurality feed on time-to-travel to urban provinces
datap = datap %>% mutate(clase1 = case_when(
  grupo == 'Urbana' ~ '',
  is.infinite(nearest.time)  ~ 'Aislada', # Isolated
  nearest.time <= 2.03 & grupo != 'Urbana' ~ 'Periurbana', # Peri-urban
  nearest.time > 2.03 & nearest.time <= 4 & grupo == 'Intermedia' ~ 'Cercana', # (Pending)
  nearest.time > 2.03 & nearest.time <= 4 & grupo == 'Rural' ~ 'Cercana a Urbana', # Close to urban
  nearest.time > 4 & grupo == 'Intermedia' ~ 'Alejada', # Far away
  nearest.time > 4 & grupo == 'Rural' ~ 'X', # (Pending)
))

datag$clase1 = datap$clase1

###########################################################################################
###########################################################################################
# DISTANCE AND TIME-TO-TRAVEL TO INTERMEDIATE PROVINCES
###########################################################################################
###########################################################################################

# Identify intermediate provinces and rural without a detailed category
int = datag %>% filter(clase1=='Cercana') %>% pull(origin)
rur = datag %>% filter(clase1=='X') %>% pull(origin)

###########################################################################################
# MAKE OPTIMAL SIZED QUERIES (TIME TO TRAVEL TO INTERMEDIATE PROVINCES)
###########################################################################################

# There are 13 intermediate (cercanas) provinces and 35 rural (X) provinces.
# We need 5 queries against 7 origins and a remaining query o keep query size below 100 elements
requ1 = list()
for(k in 1:5){
  requ[[k]] = gmapsdistance(rur[(1+7*(k-1)):(7*k)], 
                            destination = int, mode = "driving")
  print(k)
  Sys.sleep(60) # Make a query past one minute, to not overload the API
}

# Extract all the distances and time-to-travel from queries' replies
timeInt = requ[[1]]$Time
distanceInt = requ[[1]]$Distance
for(k in 2:5){
  timeInt = rbind.data.frame(timeInt, requ[[k]]$Time)
  distanceInt = rbind.data.frame(distanceInt, requ[[k]]$Distance)
}

# Save data
write.csv2(timeInt,'time_intermedias2.csv')
write.csv2(distanceInt,'distance_intermedias2.csv')

###########################################################################################
# FEED A MORE DETAILED SET OF URBAN/RURAL CATEGORIES
###########################################################################################

datag = datag %>% # Identify to which province each 'nearest' intermediate setting belongs
  left_join(timeInt %>% melt() %>% group_by(or) %>% summarise(nearint = min(value,na.rm = T)) %>%
              left_join(timeInt %>% melt(), c('or'='or','nearint'='value')), c('origin'='or'))

# Turn time-to-travel from being mesured in seconds to hours  
datap$nearint.time = datag$nearint/(60^2)   
# Extract the name of the 'capital' of nearest urban province
datap$nearest.int = gsub('.{1}$', '',str_match(datag$variable.y, "Time.\\s*(.*?)\\s*+Colombia")[,2])

# More detailed categories of rurality feed on time-to-travel to intermediate provinces
datap = datap %>% mutate(clase2 = case_when(
  is.na(nearint.time) ~ clase1, # Keep previous categories
  nearint.time <= 4 ~ 'Cercana a Intermedia', # Close to intermediate
  nearint.time > 4 ~ 'Alejada', # Far away
))

# Save complete categories
datag$clase1 = datap$clase1
write.csv(datap %>% ungroup,'clasificacion_completa2.csv')

```

## 5. Spatial network configuration of Colombian provinces
```{r}
# iGraph package for network analysis
library('igraph')

#############
# INPUT DATA
#############

# This matrix contains the distance of every province to every other one in terms of 
# time to travel (in sec.) by principal road form largest seat to largest seat: t(i,j)
# It is not defined by pairs of provinces on the same group (urban, intermediate, rural)
matriz.t = read.csv2('data/graph-matrix-distance.csv',header=T)

# This matrix contains the absolute difference of every province to every other one 
# in terms of their Demographic Urbanization Index: abs(IDUP_i - IDUP_j)
# It is not defined by pairs of provinces on the same group (urban, intermediate, rural)
matriz.f = read.csv2('data/graph-matrix-difference.csv',header=T)

# There are provinces that are isolated by road, or > 7h to a urban/intermediate city
exclude = c(16,70,72,95,124,133,138,140,145,147,148,76,105,118,150)

# Excluding isolated provinces from distance matrices
m.t = matriz.t[-exclude,-(8+exclude)]
m.t[is.na(m.t)] = 0
m.f = matriz.f[-exclude,-(8+exclude)]
m.f[is.na(m.f)] = 0

##########################
# CENTRALITY MEASURES
##########################

# Graph 1 in which weights are given by spatial distances
ig.t = graph.adjacency(as.matrix(m.t[,9:143]), mode="undirected", weighted=TRUE)
# Graph 2 in which weights are given by differences in IDUP
ig.f = graph.adjacency(as.matrix(m.f[,9:143]), mode="undirected", weighted=TRUE)

# Data frame with node-wise information
node_data = m.t[,1:8]

# Centrality #0 -- Degree centrality
dgr = degree(
  ig.t,
  v = V(ig.t),
  loops = FALSE,
  normalized = FALSE)

node_data$degree = dgr

# The rural-urban gradient is defined as the absolute difference on IDUP divided 
# over the time to travel (in hours) between the two largest municipal seats
# It can be interpreted as the force of interaction between to provinces
gradient = (E(ig.f)$weight)/(E(ig.t)$weight/(60^2))

# Centrality #1 -- Closeness centrality (in 0-100 scale)
clo = closeness(
  ig.t,
  vids = V(ig.t),
  normalized = TRUE,
  weights = 1/gradient # The weights here are interpreted as costs or distances
)

node_data$closeness.scale = 100*(clo-min(clo))/(max(clo)-min(clo))

# Centrality #2 -- Betweenness centrality (in 0-100 scale)
bet = betweenness(
  ig.t,
  v = V(ig.t),
  directed = FALSE,
  normalized = TRUE,
  weights = 1/gradient # The weights here are interpreted as costs or distances
)

node_data$betweenness.scale = 100*(bet-min(bet))/(max(bet)-min(bet))

# Centrality #3 -- Eigenvalue centrality (in 0-100 scale)
eig = eigen_centrality(
  ig.t,
  scale = TRUE,
  options = arpack_defaults,
  weights = gradient # The weights here are interpreted as forces
)

node_data$eigen.scale = 100*(eig$vector-min(eig$vector))/(max(eig$vector)-min(eig$vector))

# A glimpse on centrality measures of provinces
head(node_data)

node_data %>% 
  ggplot(aes(x=closeness.scale, y=betweenness.scale,
             color=grupo, size=eigen.scale)) +
  geom_point(alpha=0.75) +
  scale_y_sqrt() + theme_bw() +
  labs(x='Closeness centrality (0-100)',y='Betweenness centrality (0-100)') +
  guides(color=guide_legend(title="Rurality group"),
         size=guide_legend(title='Eigen centrality')) + 
  theme(legend.position='bottom')
```

```{r, fig.width = 10, fig.height=10}
##########################
# NETWORK PLOTS
##########################

# Geographic position of Colombian municipalities
geo = read.csv2('data/geo_position.csv',header=T) %>% 
  dplyr::select(codmun,La,Lo)

# Add geographic position of largest municipal seat to data
node_data = node_data %>% left_join(geo,'codmun')

# Network layout accordint to geo position
lay = norm_coords(cbind(node_data$Lo,node_data$La),
                  ymin=-1, ymax=1, xmin=-1, xmax=1)

# Assign a color depening on rurality group
colrs <- c('#b8d86c','#6cb8d8','#d86cb8')
colo = c()
for(i in 1:135){
  colo[i] = ifelse(node_data$grupo[i]=='Urbana',colrs[1],
                   ifelse(node_data$grupo[i]=='Rural',colrs[3],colrs[2]))
  }

# Adapt node labels
labels = gsub("\\.", " ", V(ig.t)$name)
labels = gsub("1", "", labels)
labels[1] = '       Bogotá'
labels[69] = 'Tumaco'

# Plot network
par(omi=c(0,0,0,0), mgp=c(0,0,0),mar=c(0,0,0,0) , family = "D")
par(mfrow=c(1,1),cex=1,cex.lab = 0.5,cex.main=0.2,cex.axis=0.2)
plot(ig.t, 
     edge.arrow.size=0,
     edge.color= 'dimgray',
     vertex.size=sqrt(node_data$IPU)/3, # Radius of node proportional to IDUP
     edge.width=(gradient)^0.8/18,	# Width is a function of gradient
     layout=lay,
     rescale=FALSE,
     vertex.shape="circle",
     vertex.label=labels,
     vertex.label.cex=0.45,
     vertex.label.dist=0.3,
     vertex.label.degree=-pi/2,
     vertex.color=colo,
     vertex.frame.color=colo,
     vertex.label.family="Helvetica")

# Plot network in circular layout
la = layout.circle(ig.t)
x = la[,1]*1.1
y = la[,2]*1.1
angle = ifelse(atan(-(la[,1]/la[,2]))*(180/pi) < 0, 
               90 + atan(-(la[,1]/la[,2]))*(180/pi),
               270 + atan(-la[,1]/la[,2])*(180/pi))
labels[1] = 'Bogotá'

par(omi=c(0,0,0,0), mgp=c(0,0,0),mar=c(0,2,0,2) , family = "D")
par(mfrow=c(1,1),cex=1,cex.lab = 0.5,cex.main=0.2,cex.axis=0.2)
plot(ig.t, 
     edge.arrow.size=0,
     edge.color= 'dimgray',
     vertex.size=sqrt(node_data$IPU)/3, # Radius of node proportional to IDUP
     edge.width=(gradient)^0.8/18,	# Width is a function of gradient
     layout=la,
     rescale=FALSE,
     vertex.shape="circle",
     vertex.label=NA,
     vertex.label.cex=NA,
     vertex.label.dist=0.5,
     vertex.label.degree=NA,
     vertex.color=colo,
     vertex.frame.color=colo)
for (i in 1:length(x)) {
  text(x=x[i], y=y[i], labels=labels[i], adj=NULL, pos=NULL, cex=.5, srt=angle[i],
       xpd=T, family = "Helvetica")
}
```
